Versioned data: why it is needed and how it can be achieved (easily and cheaply)

### 2017-09-22 ms.tex
moving random paragraphs to discussion
Minor fixes to some citations

As is evidenced in the advent of this journal, publication of quality datasets is now considered a first-class scientific product. Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013,Editorial-2014}. Datasets are now routinely archived as part of the publishing process. Moreover, increasing numbers of standalone ``Data papers'' (or descriptors) have been appearing in standard domain-level journals, as well as specialised data journals, like \emph{Scientific Data}. Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}. In particular, the current data publishing model has not yet embraced the idea that many  datasets are designed to answer scientific questions that extend beyond the scope of a single empirical paper.  These datasets are constantly evolving, i.e. never "finished", and may be the subject of repeated analysis or meta-analysis.

Many, and perhaps most, of the datasets published in this journal are likely to be living entities, meaning the state-of-the-art version of the dataset will changes through time. Typical changes may involve: the addition of new data, improving the quality of existing data, better linked to external databases, or re-structuring of the database content to be able to address new questions. For example a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names. As research around a data product grows, there might be many such additions. Ideally, any changes to a dataset would become immediately available to any interested parties, while -- to ensure reproducibility -- past versions of the dataset would remain accessible.

@@

In brief, the workflow we present, which we call Versioned Data Delivery (\textsc{vdd}), borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data. Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer. With a \textsc{vdd} system, data developers similarly maintain a core set of files (the ``code''), which produce an organised dataset that can be ``installed'' on a user's local computer. In either the development of software or data, successive versions are released over time (called ``releases''). The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data. Importantly, our \textsc{vdd} systems uses well-established tools, ensuring high-level performance and stability.

@@

% So, git doesn't actually work this way, though everyone seems to
% think that it does.  Git stores *content* and then later tries to
% de-duplicate it.  It then displays changes as line-by-line diffs
% (but you can change this to display the changes other ways; see
% --color-words --ignore-all-space for example.  The problem is that
% the algorithms that git uses are optimised for typical sizes of code
% and not for data.  Small data is actually totally fine to store with
% git.  The other issue that turns up is if the files are _binary_
% then an arbitrarily small change in the underlying data can result
% in an arbitrarily large change in the binary representation, which
% prevents deduplication.
%
% Git-LFS tries to step around this by storing large files separately
% from git's general machinery.  Git-annex did the same.  But they
% suffer from some of the other issues with git which are that it's a
% bit of a pain for (especially novice) users to do nontrivial things
% with.  I _always_ have to look up how to compare two versions across
% branches.  And there's no simple way of loading in two versions of
% the data into R without ending up with two working copies or doing
% some nasty "git show" call.
Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \cite{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
% Dan Noble: I think this is a pragmatic view, but I would have a tendency not to discuss how the tools you can use for a VDD approach are likely to be "replaced in the near future”. Of course, things change, but this will not entice the people you want the paper to appeal to to adopt this sort of approach. From my experience, some of the resistance to adopting this stuff comes from programs or repositories being “replaced” at such a high rate. They just don’t feel like its worth doing.
There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently\citep{Fli, Dat}.
% Add CKAN here?
That said, the workflow we describe here currently works well with \texttt{git} and should work even more quickly with its future replacement.

\subsection{Cloud platforms for storing and distributing version controlled data}

\subsection{Potential for collaboration: allowing for multiple data contributors}

@@

A point-and-click workflow may create barriers for reproducibility \cite{Wilson-2014,Lowndes-2017}. Ideally, users should be able to access all versions of the database programmatically. Code to access the \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle for reproducibility.  To make access via code as easy as possible for users, we offer a novel implementation of a \textsc{vdd} system. We have written an \textsc{R} package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the \texttt{GitHub} \textsc{api} to pull down versions of the dataset; and b) constructs the shell of a second, database specific \textsc{R} package, which can be used to access versions of the dataset. Using this system, a researcher can create and distribute a \textsc{R} package that facilitates access to their data with (very) minimal computational skills. We have focussed on the \texttt{R} platform \cite{R-2017} as one of the most prominent platforms for data analysis.

For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \cite{Falster-2015} stored at \smurl{github.com/dfalster/baad}. \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}. And importantly \texttt{baad.data} contains no actual data so is very quick to install and takes up virtually no space on the user's hard-drive.
% Dan Noble - Next 5 lines: I found this part to be a bit difficult to grasp as you discuss how updated versions can "download the latest version”, but then in the same sentence discuss how this works seamlessly “offline”. You may want to discuss this in greater depth as I don’t see how this is possible.
When the function \texttt{baad.data::baad\_data()} is called, the system will download the latest version; if there have not been any changes since the last time the function was called, it will simply retrieve the data from memory, making the system very efficient (and also allowing the system to work seamlessly if the computer is offline). If a researcher requires an older version of \textsc{BAAD}, specifying the version number of the data as an argument, e.g. \texttt{baad.data::baad\_data("1.0.1")} will retrieve it, without overwriting any other dataset versions they may have previously downloaded. This enables users to reproduce an analysis using a specific version of the data, even while new data is posted.

@@

>>>>

As is evidenced in the advent of this journal \cite{Editorial-2014}, publication of quality datasets is now considered a first-class scientific product. Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}. Datasets are now routinely archived as part of the publishing process. Moreover, increasing numbers of standalone ``Data papers'' (or descriptors) have been appearing in standard domain-level journals, as well as specialised data journals. Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \citep{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}. In particular, the current data publishing model has not yet embraced the idea that many  datasets are designed to answer scientific questions that extend beyond the scope of a single empirical paper.  These datasets are constantly evolving, i.e. never "finished", and may be the subject of repeated analysis or meta-analysis.

Many, and perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art version of the dataset will changes through time. Typical changes may involve: the addition of new data, improving the quality of existing data, better linked to external databases, or re-structuring of the database content to be able to address new questions. For example a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names. As research around a data product grows, there might be many such additions. Ideally, any changes to a dataset would become immediately available to any interested parties, while -- to ensure reproducibility -- past versions of the dataset would remain accessible.

@@

In brief, the workflow we present, which we call Versioned Data Delivery (\textsc{vdd}), borrows best-practices for software development \citep{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data. Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer. With a \textsc{vdd} system, data developers similarly maintain a core set of files (the ``code''), which produce an organised dataset that can be ``installed'' on a user's local computer. In either the development of software or data, successive versions are released over time (called ``releases''). The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data. Importantly, our \textsc{vdd} systems uses well-established tools, ensuring high-level performance and stability.

@@

% So, git doesn't actually work this way, though everyone seems to
% think that it does.  Git stores *content* and then later tries to
% de-duplicate it.  It then displays changes as line-by-line diffs
% (but you can change this to display the changes other ways; see
% --color-words --ignore-all-space for example.  The problem is that
% the algorithms that git uses are optimised for typical sizes of code
% and not for data.  Small data is actually totally fine to store with
% git.  The other issue that turns up is if the files are _binary_
% then an arbitrarily small change in the underlying data can result
% in an arbitrarily large change in the binary representation, which
% prevents deduplication.
%
% Git-LFS tries to step around this by storing large files separately
% from git's general machinery.  Git-annex did the same.  But they
% suffer from some of the other issues with git which are that it's a
% bit of a pain for (especially novice) users to do nontrivial things
% with.  I _always_ have to look up how to compare two versions across
% branches.  And there's no simple way of loading in two versions of
% the data into R without ending up with two working copies or doing
% some nasty "git show" call.
Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \citep{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
% Dan Noble: I think this is a pragmatic view, but I would have a tendency not to discuss how the tools you can use for a VDD approach are likely to be "replaced in the near future”. Of course, things change, but this will not entice the people you want the paper to appeal to to adopt this sort of approach. From my experience, some of the resistance to adopting this stuff comes from programs or repositories being “replaced” at such a high rate. They just don’t feel like its worth doing.
There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently\citep{Fli, Dat}.
% Add CKAN here?
That said, the workflow we describe here currently works well with \texttt{git} and should work even more quickly with its future replacement.

@@

A point-and-click workflow may create barriers for reproducibility \citep{Wilson-2014,Lowndes-2017}. Ideally, users should be able to access all versions of the database programmatically. Code to access the \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle for reproducibility.  To make access via code as easy as possible for users, we offer a novel implementation of a \textsc{vdd} system. We have written an \textsc{R} package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the \texttt{GitHub} \textsc{api} to pull down versions of the dataset; and b) constructs the shell of a second, database specific \textsc{R} package, which can be used to access versions of the dataset. Using this system, a researcher can create and distribute a \textsc{R} package that facilitates access to their data with (very) minimal computational skills. We have focussed on the \texttt{R} platform \cite{R-2017} as one of the most prominent platforms for data analysis.

For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \citep{Falster-2015} stored at \smurl{github.com/dfalster/baad}. \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}. And importantly \texttt{baad.data} contains no actual data so is very quick to install and takes up virtually no space on the user's hard-drive.
% Dan Noble - Next 5 lines: I found this part to be a bit difficult to grasp as you discuss how updated versions can "download the latest version”, but then in the same sentence discuss how this works seamlessly “offline”. You may want to discuss this in greater depth as I don’t see how this is possible.
When the function \texttt{baad.data::baad\_data()} is called, the system will download the latest version; if there have not been any changes since the last time the function was called, it will simply retrieve the data from memory, making the system very efficient (and also allowing the system to work seamlessly if the computer is offline). If a researcher requires an older version of \textsc{BAAD}, specifying the version number of the data as an argument, e.g. \texttt{baad.data::baad\_data("1.0.1")} will retrieve it, without overwriting any other dataset versions they may have previously downloaded. This enables users to reproduce an analysis using a specific version of the data, even while new data is posted.

@@

% potentially move to discussion? 
\subsection{Cloud platforms for storing and distributing version controlled data}

% potentially move to discussion?
\subsection{Potential for collaboration: allowing for multiple data contributors}
